Practical Optional Types for Clojure

### 2015-02-25 overview.tex
fixes from my office
hmap examples

Clojure~\cite{Hic08} is a Lisp built to run on the
Java Virtual Machine with exemplary support for concurrent programming
and immutable data structures. It emphaisizes mostly-functional
programming, restricting imperative updates to a limited set of
structures which have specific thread synchronization behavior. By
default, it provides fast implemenations of immutable lists, vectors,
and hash tables, which are used for most data structures, although it
also provides means for defining new records.

@@

Classic occurrence typing concentrates on conditional
control flow like the following example.

\begin{listing}
\inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
\caption{Occurrence typing}
\end{listing}

Here we include the full \clj{ns} form, which we omit in the following
examples. Going down the then branch in this example means that we know
\clj{x} is a number, which is automatically updated with occurrence typing.

In Clojure, exceptional control flow is also common, often in the form of
runtime assertions.

\begin{listing}
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/do.clj}
\caption{Exceptional control flow}
\end{listing}

Clojure assertions expand to simple exceptional control flow,
which we enhance our sequencing \doliteral{} expression to understand.

\begin{minted}{clojure}
(fn [x :- (U nil Number)]
  (do (if (number? x) nil (throw (new Exception)))
      (inc x)))
\end{minted}

Our main insight is that occurrence typing already provides
valuable information about the left argument of \doliteral{} expressions,
namely if it returns a value, \clj{x} is a \clj{Number}.
Since the right argument is evaluated after the left, we can assume
\clj{x} is a \clj{Number} when evaluating \clj{(inc x)}, verifying
the programmer's attempt to avoid null-pointer exceptions without further annotations.

@@

Hash-maps with keyword keys play a major role in Clojure programming.

HMaps track the presense and absense of entries.

$$
\clj{(HMap :mandatory {:a Num})} 
$$

is the type of a map with at least an \clj{:a} entry mapped to a number,
also written \clj{'{:a Num}}. If \clj{:b} is also known to be an absent entry
then

% '#' not allowed in mintinline
\smallskip
\begin{minted}{clojure}
  (HMap :mandatory {:a Num} :absent-keys #{:b})
\end{minted}

declares \clj{:b} absent. If the mandatory keys fully specify the
entries, we write

$$
\clj{(HMap :mandatory {:a Num} :complete? true)}.
$$

\clj{:complete?} defaults to \clj{false}.
The \clj{:optional} option

$$
\clj{(HMap :optional {:a Num})}
$$

abbreviates the union of present and absent entries

\smallskip
\begin{minted}{clojure}
(U '{:a Num} (HMap :absent #{:a}))
\end{minted}
.

\clj{defalias} defines a custom type abbreviation, often for unions of HMaps.

\begin{listing}
\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
\inputminted[firstline=26,lastline=28]{clojure}{code/demo/src/demo/hmap.clj}
\caption{A simple AST in terms of HMaps}
\end{listing}

HMaps are very malleable, including support for associate, merge and dissociate operations.

\begin{listing}
\inputminted[firstline=10,lastline=24]{clojure}{code/demo/src/demo/hmap.clj}
\caption{HMap operations}
\end{listing}

Lookups on HMaps with multiple options combined with
occurrence typing is an interesting problem.
Testing a lookup on a mandatory entry is straightforward, since we
know the lookup always succeeded.

\begin{minted}{clojure}
(fn [m :- '{:a (U nil Num)}] :- '{:a Num}
  {:pre [(number? (:a m))]}
  m)
\end{minted}

The result type is updated as expected to exclude \clj{nil}.
Testing for \clj{nil} is more complicated, since it is the sentinel
lookup failure indicator but may also appear as an entry value.
If we test for \clj{nil} in an \clj{:optional} entry, the best we learn
is that either the entry is missing or contains \clj{nil}.

\begin{minted}{clojure}
(fn [m :- (HMap :optional {:a (U nil Num)})] 
  :- (HMap :optional {:a nil})
  {:pre [(nil? (:a m))]}
  m)
\end{minted}

>>>>

Clojure~\cite{Hic08} is a Lisp built to run on the
Java Virtual Machine with exemplary support for concurrent programming
and immutable data structures. It emphasizes mostly-functional
programming, restricting imperative updates to a limited set of
structures which have specific thread synchronization behaviour. By
default, it provides fast implementations of immutable lists, vectors,
and hash tables, which are used for most data structures, although it
also provides means for defining new records.

@@

Tobin-Hochstadt and Felleisen~\cite{TF10}
introduced Typed Racket with occurrence typing,
an inference technique for conditional control flow.
They introduce the concept of occurrence typing 
with the following example.

\inputminted[firstline=1]{racket}{code/tr/example1.rkt}

This function takes a value that is either \emph{\#f} % mintinline really hates #
or a number, represented by an untagged \emph{union type}.
The `then' branch has an implicit invariant
that \rkt{x} is a number, which is automatically inferred with occurrence typing
and type checked without further annotations.
Here is same program in Typed Clojure.

\begin{exmp}
\inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
\end{exmp}

Clojure uses \emph{namespaces}, declared with the
\clj{ns} form, to manage top-level \emph{vars}.
This is a regular Clojure file compiled with
the Clojure compiler except it declares a runtime dependency on
\clj{clojure.core.typed}, Typed Clojure's core namespace.
Programmers import typed versions of certain constructs as needed, like
a \clj{fn} variant that support annotations,
and use the provided \clj{check-ns} function to type check the current namespace
when convenient.

Typed Clojure can check all the examples in the occurrence typing
paper---the rest of this section describes the extensions necessary
to check Clojure code.

\subsection{Exceptional control flow}

Along with conditional control flow,
Clojure programmers rely on \emph{exceptions}
to assert type-related invariants.

\begin{exmp}
\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
\end{exmp}

We combine the sequencing form \clj{do} and a thrown exception with
\clj{throw} to avoid a possible null-pointer exception by incrementing \clj{nil}.
(We omit \clj{ns} forms for the rest of the examples).

To check this example, we note that
occurrence typing already provides
valuable information about the conditional expression---namely that it cannot return logically true and if it returns logically false 
then \clj{x} is a \clj{Number}. Equivalently, if it returns a value then \clj{x} is a \clj{Number}.
We use this assumption to check \clj{(inc x)}, automatically
ruling out the possibility of null-pointer exceptions
(said formally in section~\ref{sec:doformal}).

@@

Hash-maps with keyword keys play a major role in Clojure programming.
HMap types model the most common usages of keyword maps.

\begin{exmp}
\begin{listing}
\inputminted[firstline=5,lastline=8]{clojure}{code/demo/src/demo/hmap.clj}
\inputminted[firstline=26,lastline=28]{clojure}{code/demo/src/demo/hmap.clj}
\end{listing}
\end{exmp}

Here we define a type abbreviation with \clj{defalias} called \clj{Expr}
that describes the structure of a recursively-defined AST as a union of HMaps.
\clj{an-exp} is a function verified to return an \clj{Expr}.

HMaps support various options to track the presence and absence of entries.
\clj{'{:a Num}} is a shorthand for the \clj{:mandatory} option---entries that
must be present:
$$
\clj{(HMap :mandatory {:a Num})}.
$$

\clj{:absent-keys} is a set of keywords that are known to not have entries.
The value \clj{{:a 1 :b nil}} is not a member of the type
% '#' not allowed in mintinline
\smallskip
\begin{minted}{clojure}
  (HMap :mandatory {:a Num} :absent-keys #{:b}).
\end{minted}

\clj{:optional} is a map that declares entries either absent or
present with a specific type. Concretely
$$
\clj{(HMap :optional {:a Num})}
$$
is the same as
\begin{minted}{clojure}
      (U '{:a Num} (HMap :absent-keys #{:a})).
\end{minted}

\clj{:complete?} is \clj{true} if there are no further entries
than those declared \clj{:mandatory} or \clj{:optional}.

\clj{:mandatory}, \clj{:optional} and \clj{:absent-keys} default to empty
and \clj{:complete?} defaults to \clj{false}.

HMaps interact with occurrence typing in interesting ways.
What do we learn about the map \clj{m} in the following test?

\begin{exmp}
\inputminted[firstline=5,lastline=9]{racket}{code/demo/src/demo/hmap_path.clj}
\end{exmp}

In the then branch Typed Clojure infers \clj{'{:a Num}}, in the else branch
\clj{(HMap :optional {:a nil})}. Failed lookups return \clj{nil},
so we cannot distinguish between \clj{'{:a nil}} and \clj{(HMap :optional {:a nil})}
by testing the result of looking up \clj{:a}.
Section~\ref{sec:hmapformal} discusses this in greater detail.

### 2015-02-25 small.tex
tweak intro example. Ack.
add appendix
improve intro
fixes from my office
hmap examples

\usepackage{common}
\usepackage{minted}
\include{bibliography.bib}

@@

@@

\begin{abstract}
Typed Clojure is an optional type system for Clojure, a dynamic
language in the Lisp family that targets the JVM. Typed Clojure's type
system build on the design of Typed Racket, repurposing in particular
\emph{occurence typing}, its approach to statically reasoning about
predicate tests. However, in adapting the type system to Clojure,
changes and extensions are required to accomodate both additional
language features and  idioms used by Clojure programmers.

In this paper, we describe Typed Clojure and present these type system
extensions, focusing on three features widely used in Clojure. First, Java
interoperability is central to Clojure's mission but introduces
challenges such as ubiquitous \texttt{null}; Typed Clojure handles
Java interoperability while ensuring the absence of null-pointer
exceptions in typed programs. Second, Clojure programmers
idiomatically use immutable dictionaries for data structures; Typed
Clojure handles this in the type system with multiple forms of
heterogenous dictionary types. Third, multimethods provide extensible
operations, and their Clojure semantics turns out to have a surprising
synergy with the underlying occurence typing framework.

@@

@@

>>>>

\usepackage{common}
\usepackage{minted}
\include{bibliography.bib}
\usepackage{chngcntr}

\newcounter{ex}
\newenvironment{exmp}{
\refstepcounter{ex}%
\par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
\\[-2.5em]\nopagebreak}
{\relax}

@@

\newcommand{\clj}[1]{\mintinline{clojure}{#1}}
\newcommand{\rkt}[1]{\mintinline{racket}{#1}}

@@

\begin{abstract}
Typed Clojure is an optional type system for Clojure, a dynamic
language in the Lisp family that targets the JVM. Typed Clojure's type
system build on the design of Typed Racket, repurposing in particular
\emph{occurrence typing}, its approach to statically reasoning about
predicate tests. However, in adapting the type system to Clojure,
changes and extensions are required to accommodate both additional
language features and  idioms used by Clojure programmers.

In this paper, we describe Typed Clojure and present these type system
extensions, focusing on three features widely used in Clojure. First, Java
interoperability is central to Clojure's mission but introduces
challenges such as ubiquitous \texttt{null}; Typed Clojure handles
Java interoperability while ensuring the absence of null-pointer
exceptions in typed programs. Second, Clojure programmers
idiomatically use immutable dictionaries for data structures; Typed
Clojure handles this in the type system with multiple forms of
heterogenous dictionary types. Third, multimethods provide extensible
operations, and their Clojure semantics turns out to have a surprising
synergy with the underlying occurrence typing framework.

@@

\input{acknowledgements}

@@

\clearpage

\input{appendix}
